---
title: "Tidy Tuesday: NHL Goalscorers"
author: "Jonny Law"
date: "2020-03-03"
slug: tidy-tuesday/
aliases: [tidy-tuesday]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggrepel)
library(tidytuesdayR)
```

First install the Tidy Tuesday R package.

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("thebioengineer/tidytuesdayR")
```

The data for this Tuesday can be downloaded using `tt_load`.

```{r}
tuesdata <- tidytuesdayR::tt_load('2020-03-03')
```

Unfortunately this only grabbed one file - the top 250 goalscorers. First look at this file.

# Top Career Goalscorers

```{r}
tuesdata$top_250 %>% 
  top_n(30, wt = total_goals) %>% 
  mutate(player = forcats::fct_reorder(player, total_goals)) %>% 
  ggplot(aes(x = player, y = total_goals)) +
  geom_col() +
  coord_flip() 
```

# Average goals per season

```{r}
parse_end_year <- function(years) {
  end_tens <- substr(years, 6, 7)
  possible_end <- as.numeric(paste0(substr(years, 1, 2), end_tens))
  start <- as.numeric(substr(years, 1, 4))
  if (possible_end - start > 0) {
    possible_end
  } else {
    as.numeric(paste0(20, end_tens))
  }
}

options(scipen = 99)
tuesdata$top_250 %>%
  rowwise() %>% 
  mutate(
    yr_end = parse_end_year(years),
    seasons = yr_end - yr_start,
    average_goals_per_season = total_goals / seasons,
    decade = cut(yr_start, breaks = seq(1920, 2020, by = 10), dig.lab = 10)
  ) %>% 
  ungroup() %>% 
  # group_by(decade) %>% 
  mutate(player = forcats::fct_reorder(player, average_goals_per_season)) %>% 
  top_n(30, wt = average_goals_per_season) %>%
  ggplot(aes(x = player, y = average_goals_per_season, fill = active)) +
  geom_col() +
  coord_flip()
  # facet_wrap( ~ decade, scales = "free_y")
```

```{r}
goals_per_season <- tuesdata$top_250 %>%
  rowwise() %>%
  mutate(
    yr_end = parse_end_year(years),
    seasons = yr_end - yr_start,
    average_goals_per_season = total_goals / seasons,
    decade = cut(
      yr_start,
      breaks = seq(1920, 2020, by = 10),
      dig.lab = 10
    )
  ) %>%
  ungroup()

ggplot() +
  geom_point(data = goals_per_season,
             aes(x = seasons, y = average_goals_per_season)) +
  ggrepel::geom_label_repel(data = top_n(goals_per_season, 10, total_goals), aes(x = seasons, y = average_goals_per_season, label = paste(player, total_goals), fill = active)) +
  xlab("Total Seasons in NHL") +
  ylab("Average Goals per Season") +
  theme_minimal() +
  labs(title = tools::toTitleCase("average goals per season and total number of seasons \nfor the top 250 NHL Goal Scorers"), subtitle = "Top 10 all-time goal scorers are labelled") +
  theme(legend.position = c(0.9, 0.9), legend.title = element_blank())

# ggsave("goals_per_season.png")
```

# Game Goals

```{r}
game_goals <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-03/game_goals.csv')
```

```{r}
highlighted_players <- tuesdata$top_250 %>% 
  filter(player %in% c("Alex Ovechkin", "Wayne Gretzky")) %>% 
  select(player, yr_start)

cumulative_goals <- game_goals %>% 
  mutate(age_years = as.numeric(substr(age, 1, 2))) %>% 
  group_by(player, age_years) %>% 
  summarise(goals_per_season = sum(goals)) %>% 
  arrange(player, age_years) %>% 
  inner_join(tuesdata$top_250, by = "player") %>% 
  mutate(cumulative_goals = cumsum(goals_per_season))

ggplot(cumulative_goals, aes(x = age_years, y = cumulative_goals)) +
  geom_line(data = cumulative_goals,
            aes(x = age_years, y = cumulative_goals, group = player),
            alpha = 0.2) +
  geom_line(
    data = cumulative_goals %>% inner_join(highlighted_players, by = "player"),
    aes(x = age_years, y = cumulative_goals, colour = player)
  ) +
  geom_label_repel(
    data =
      cumulative_goals %>%
      inner_join(highlighted_players, by = "player") %>% 
      filter(age_years == 30),
    aes(
      x = age_years,
      y = cumulative_goals,
      label = player,
      colour = player
    )
  ) +
  xlab("Player Age") +
  ylab("Total Goals") +
  theme_minimal() +
  labs(title = "Cumulative Goals in the NHL by Age", subtitle = "Data from 42 of the all time 250 NHL scorers who started their career since game-level \ndata became available in 1979/80 season.") +
  theme(legend.position = "none")

# ggsave("cumulative_goals.png")
```

## Post some results to Twitter

Now use the [twitteR](https://cran.r-project.org/web/packages/twitteR/README.html) library to post the plots directly to Twitter without leaving R. You must connect to the Twitter API using OAuth as described in the README for twitteR, I set the Twitter application keys in my `.Renviron` file which is never committed to public version control (this can be easily edited using `usethis::edit_r_environ()`).

```{r eval=FALSE}
library(twitteR)
oauth <- setup_twitter_oauth(
  Sys.getenv("TWITTER_API_KEY"),
  Sys.getenv("TWITTER_API_SECRET_KEY"),
  access_token = Sys.getenv("TWITTER_ACCESS_TOKEN"),
  access_secret = Sys.getenv("TWITTER_ACCESS_TOKEN_SECRET")
)
```

With twitteR we can get the most recent #tidytuesday posts

```{r eval=FALSE}
searchTwitter('#tidytuesday', n = 10)
```

Unfortunately the twitteR API (I think people tend to use [rtweet](https://rtweet.info) now!) is a little outdated and requires tweets to be less than 140 characters (however this can be bypassed using `bypassCharLimit = TRUE`) and we can't post multiple images in the same tweet using the `mediaPath` argument. So unfortunately I had to resort to posting my graphs manually!

```{r eval=FALSE}
updateStatus(
  "Can Alex Ovechkin overhaul Wayne Gretzky's all time NHL goal scoring record. It appears as if Wayne slowed down in the latter years of his career. #rstats #TidyTuesday",
  mediaPath = c("cumulative_goals.png", "goals_per_season.png"),
  bypassCharLimit = TRUE
)
```

