---
title: "Gibbs Sampling for the Hidden Markov Model"
description: |
  A short description of the post.
author:
  - name: Jonny Law
    url: {}
date: 07-30-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(jonnylaw)
```

Calculate $p(z_t = i | z_{t+1} = j, y_{1:T})$

```{r}
hmm_backward_sample_step <- function(
    sampled_state,
    alpha,
    alpha1,
    y1,
    transition_matrix,
    observation) {
    observation_dist = observation(y1, sampledState)
    transition = transitionMatrix[, sampledState]

    probs = ((observation_dist %*% transition) * alpha) / alpha1

    sample(x = seq_len(length(probs)), size = 1, prob = probs)
  }
```

Perform backward sampling for the HMM

```{r}
hmm_backward_sample <- function(ys,
                                alphas, # matrix
                                observation,
                                transition_matrix) {
  d <- ncol(alphas)
  n <- length(ys)
  sampleN = sample(seq_len(d), size = 1, prob = alphas[n,])
  
  purrr::accumulate(
    rev(ys),
    hmm_backward_step,
    observation = observation, 
    transition_matrix = transition_matrix, 
    .init = sampleN
  )
    
    (alphas.init).lazyZip(alphas.tail).lazyZip(ys.tail).toVector.
    scanRight(sampleN) {
      case ((alpha0, alpha1, y1), sample) =>
        backwardSampleStep(transitionMatrix, observation, alpha0, alpha1, y1, sample)
    }
  }
```

Sample a value of the latent-state of an HMM

```{r}
hmm_forward_backward_sample <- function(ys,
                                        x0,
                                        observation,
                                        transitionMatrix) = {
  alphas = hmm_forward(ys, x0, observation, transitionMatrix)
  hmm_backward_sample(ys, alphas, observation, transitionMatrix)
}
```

Sample a value of the transition matrix from a product of Dirichlet distributions

```{r}
sample_theta <- function(xs: Vector[Int], alpha: DenseVector[Double]): DenseMatrix[Double] = {
    val nStates = 2
    val stateTransitions = Ctmc.stateTransitions(nStates, xs)
    val out: Array[Array[Double]] = breeze.util.JavaArrayOps.dmDToArray2(stateTransitions).
      map(x => Dirichlet(alpha + DenseVector(x)).draw.data)
    breeze.util.JavaArrayOps.array2DToDm(out)
  }

  case class GibbsState(transitionMatrix: DenseMatrix[Double], xs: Vector[Int])
```
  /**
   * Perform a single Gibbs step for a discrete time HMM
   *
   * @param alpha
   * @param ys
   * @param x0
   * @param observation
   * @param state
   * @return
   */
    
```{r}
  def gibbsStep[A](
    alpha: DenseVector[Double],
    ys: Vector[A],
    x0: DenseVector[Double],
    observation: A => DenseVector[Double])
               (state: GibbsState) = {
    val transitionMatrix = sampleTheta(state.xs, alpha)
    val newState = forwardBackwardSample(ys, x0, observation, transitionMatrix)
    GibbsState(transitionMatrix, newState)
  }
```

  /**
   * Perform Gibbs Sampling for a discrete time HMM with known initial distribution
   * and observation matrix
   *
   * @param alpha prior for the dirichlet distribution
   * @param ys observations of the HMM
   * @param x0 initial distribution of the states
   * @param observation observation distribution
   */
    
```{r}
  def gibbsHmm[A](
    alpha: DenseVector[Double],
    ys: Vector[A],
    x0: DenseVector[Double],
    observation: A => DenseVector[Double]): Iterator[GibbsState] = {
    val nStates = x0.size
    val transitionMatrix = new DenseMatrix(nStates, nStates, Dirichlet(alpha).sample(nStates).toArray.flatMap(_.data))
    val xs = forwardBackwardSample(ys, x0, observation, transitionMatrix)
    Iterator.iterate(GibbsState(transitionMatrix, xs))(gibbsStep(alpha, ys, x0, observation))
  }
```

