---
title: "Gibbs Sampling for the Hidden Markov Model"
description: |
  A short description of the post.
author:
  - name: Jonny Law
    url: {https://twitter.com/lawsy}
date: 07-30-2020
draft: TRUE
output:
  distill::distill_article:
    self_contained: false
categories:
  - Scala
  - Bayesian
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(jonnylaw)
```

In this post we will consider the backward smoothing algorithm for a hidden Markov model (HMM). While exploring the derivation and implementation we will consider how to avoid the use of mutable state and implement the algorithm in a purely functional way. This post follows from the previous post discussing [hidden Markov models] and functional programming, the inference scheme developed in that post was a Metropolis-Hastings algorithm where the marginal likelihood calculated using the HMM forward algorithm.

In this post I am going to use the functional programming language Scala to illustrate the ideas required to implement the backward smoothing algorithm for the HMM. As an extension, we will consider another parameter inference algorithm, a Gibbs sampling algorithm targeting the posterior distribution of the transition matrix.

The 

First we sample $p(z_t = i | z_{t+1} = j, y_{1:T})$

```scala
def stateTransitions(
  nStates: Int,
  states: Vector[Int]): DenseMatrix[Double] = {
  val zeros = DenseMatrix.zeros[Double](nStates, nStates)
  states.init.zip(states.tail).foldLeft(zeros){
    case (acc, (from, to)) => acc.update(from, to, acc(from, to) + 1)
    acc
  }
}
```

```scala
 /**
   * A single step of the backward pass
   *
   * @param observation
   * @param transitionMatrix
   * @param y
   * @param beta
   */
  def backwardStep[A](
    observation: A => DenseVector[Double],
    transitionMatrix: DenseMatrix[Double])(
    y: A,
    beta: DenseVector[Double]) = {
    normalise(transitionMatrix * (observation(y) *:* beta))
  }

  /**
   * Backward pass of the forward-backward algorithm
   *
   * @param ys
   * @param transitionMatrix
   * @param observation
   * @return
   */
  def backward[A](
    ys: Vector[A],
    transitionMatrix: DenseMatrix[Double],
    observation: A => DenseVector[Double]): Vector[DenseVector[Double]] = {
    val nStates = transitionMatrix.cols
    val init = DenseVector.ones[Double](nStates)
    ys.scanRight(init)(backwardStep(observation, transitionMatrix))
  }

  /**
   * Calculate the smoothed states from the forward and backward smoothing algorithms
   */
  def smoothed(
    alphas: Vector[DenseVector[Double]],
    betas: Vector[DenseVector[Double]]): Vector[DenseVector[Double]] = {
    alphas.zip(betas).map { case (a, b) => normalise(a *:* b) }
  }
  
  /**
   * Perform Forward-Backward smoothing
   *
   * @param ys
   * @param x0
   * @param observation
   * @param transitionMatrix
   * @return
   */
  def fowardBackward[A](
    ys: Vector[A],
    x0: DenseVector[Double],
    observation: A => DenseVector[Double],
    transitionMatrix: DenseMatrix[Double]): Vector[DenseVector[Double]] = {
    val alphas = forward(ys, x0, observation, transitionMatrix)
    val betas = backward(ys, transitionMatrix, observation)
    smoothed(alphas, betas)
  }
```

## Gibbs Sampler

```scala
  /**
   * Calculate p(z_t = i | z_{t+1} = j, y_{1:T})
   *
   * @param transitionMatrix
   * @param observation
   * @param alpha
   * @param alpha1
   */
  def backwardSampleStep[A](
    transitionMatrix: DenseMatrix[Double],
    observation: A => DenseVector[Double],
    alpha: DenseVector[Double],
    alpha1: DenseVector[Double],
    y1: A,
    sampledState: Int): Int = {
    val observationDist: Double = observation(y1)(sampledState)
    val transition: DenseVector[Double] = transitionMatrix(::, sampledState)

    val probs = ((observationDist * transition) *:* alpha) /:/ alpha1

    Multinomial(probs).draw
  }

  /**
   * Perform backward sampling for the HMM
   *
   * @param ys
   * @param alphas
   * @param observation
   * @param transitionMatrix
   * @return
   */
  def backwardSample[A](
    ys: Vector[A],
    alphas: Vector[DenseVector[Double]],
    observation: A => DenseVector[Double],
    transitionMatrix: DenseMatrix[Double]): Vector[Int] = {
    val sampleN: Int = Multinomial(alphas.last).draw

    (alphas.init).lazyZip(alphas.tail).lazyZip(ys.tail).toVector.
      scanRight(sampleN) { case ((alpha0, alpha1, y1), sample) =>
        backwardSampleStep(transitionMatrix, observation, alpha0, alpha1, y1, sample) }
  }

  /**
   * Sample a value of the latent-state of an HMM
   */
  def forwardBackwardSample[A](
    ys: Vector[A],
    x0: DenseVector[Double],
    observation: A => DenseVector[Double],
    transitionMatrix: DenseMatrix[Double]): Vector[Int] = {
    val alphas = forward(ys, x0, observation, transitionMatrix)
    backwardSample(ys, alphas, observation, transitionMatrix)
  }

  /**
   * Sample a value of the transition matrix from a product of Dirichlet distributions
   */
  def sampleTheta(xs: Vector[Int], alpha: DenseVector[Double]): DenseMatrix[Double] = {
    val nStates = 2
    val stateTransitions = Ctmc.stateTransitions(nStates, xs)
    val out: Array[Array[Double]] = breeze.util.JavaArrayOps.dmDToArray2(stateTransitions).
      map(x => Dirichlet(alpha + DenseVector(x)).draw.data)
    breeze.util.JavaArrayOps.array2DToDm(out)
  }

  case class GibbsState(transitionMatrix: DenseMatrix[Double], xs: Vector[Int])
```

## Dishonest Casino

```{r}
gibbs <-
  read_csv(
    "~/git/ctmc/models/dishonest_casino_gibbs.csv",
    col_names = c("a",
                  "one_minus_a",
                  "one_minus_b",
                  "b")
  )
```


```{r}
gibbs %>% 
  select(a, b) %>% 
  mutate(iteration = row_number()) %>% 
  pivot_longer(c("a", "b"), names_to = "parameter", values_to = "value") %>% 
  jonnylaw::plot_diagnostics_sim(actual_values = 
                                   tibble(parameter = c("a", "b"), actual_value = c(0.3, 0.1)))
```

## Simple HMM

The transition matrix used to simulate the sequence of states and observations is

$$P = \begin{pmatrix}
0.5 & 0.5 \\
0.4 & 0.6
\end{pmatrix}$$

The emission matrix used to simulate the sequence of states and observations is

$$B = \begin{pmatrix}
0.2 & 0.8 \\
0.9 & 0.1
\end{pmatrix}.$$

```{r}
simple_hmm <- read_csv("~/git/ctmc/data/simple_hmm.csv")

simple_hmm %>% 
  filter(time < 100) %>% 
  pivot_longer(-time, values_to = "value", names_to = "key") %>% 
  ggplot(aes(x = time, y = value)) +
  geom_step() +
  facet_wrap(~key, ncol = 1)
```

```{r simple_hmm}
gibbs <- read_csv("~/git/ctmc/models/simple_hmm_gibbs.csv")

gibbs %>% 
  group_by(chain) %>% 
  mutate(iteration = row_number()) %>% 
  pivot_longer(c("p1", "p2", "p3", "p4"), names_to = "parameter", values_to = "value") %>% 
  jonnylaw::plot_diagnostics()
```


